<<<<<<< HEAD
---
title: "ecosense_proto_vignette"
author: "George A. Whitehouse"
date: "11/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rpath); library(data.table); library(viridis)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Rpath is an R implementation of the food web modeling program Ecopath with Ecosim
(EwE; Christensen and Pauly 1992^[Christensen V, Pauly D (1992) Ecopath II-a software 
for balancing steady-state ecosystem models and calculating network characteristics. 
Ecol Model 61:169-185. doi:10.1016/0304-3800(92)90016-8], Walters et al. 1997^[Walters 
C, Christensen V, Pauly D (1997) Structuring dynamic models of exploited ecosystems 
from trophic mass-balance assessments. Rev Fish Biol Fish 7:139-172. 
doi:10.1023/a:1018479526149]). For full documentation of the Rpath package see Lucey 
et al. (2020^[Lucey SM, Gaichas SK, Aydin KY (2020) Conducting reproducible ecosystem 
modeling using the open source mass balance model Rpath. Ecol Model 427:11. 
doi:10.1016/j.ecolmodel.2020.109057]). Ecosense is a Monte Carlo approach to generating 
an ensemble of plausible ecosystem parameter sets from a single Rpath model for 
use with `rsim.run`, the time-dynamic counterpart to Rpath. For complete documentation
of Ecosense see Whitehouse and Aydin (2020^[Whitehouse GA, Aydin KY (2020) 
Assessing the sensitivity of three Alaska marine food webs to perturbations: an 
example of Ecosim simulations using Rpath. Ecol Model 429:16. 
doi:10.1016/j.ecolmodel.2020.109074]). This vignette assumes the reader is familiar
with EwE and the Rpath package and its basic operations.

## Ecosense

To use Ecosense you need an `rpath.params` object and the corresponding `rsim.scenario`
object. In this vignette we've included the example Rpath models from Whitehouse and 
Aydin (2020) for the eastern Bering Sea (EBS), eastern Chukchi Sea (ECS), and Gulf 
of Alaska (GOA) marine food webs, and use the EBS model in our examples. Each of 
these models has 50–54 functional groups, including one fishery, one primary 
producer, two detrital compartments, and there are no stanzas.

```{r load the models and Ecosense}
load("data/sense_vignette.RData")
ls()
source("R/ecosense.R", local = knitr::knit_global())
```
Each of the three Rpath models included in this data set has an `rpath.params` object (unbal), 
a balanced `rpath` model (bal), and an `rsim.scenario` object (scene).

### Setting up Ecosense

Full sets of Ecosim parameter sets are drawn from distributions centered on the
original Rpath parameter estimates. The width of the distribution for each parameter
is defined in the data pedigree of the `rpath.params` object. The drawn parameters
include biomass, P/B, Q/B, diet composition, and natural mortality (i.e., M zero).
Additionally, vulnerability and handling time from the predator prey functional 
response can be drawn over a specified range (lower/upper bounds in log space-1) 
for each individual predator-prey interactions.

A single parameter set can be generated with a call to `rsim.sense`. The object 
returned by `rsim.sense` is equivalent to the list of parameters in an `rsim.scenario` 
object (e.g., EBS_scene$params).

```{r echo=TRUE}
# One set of Ecosim parameters for the EBS model
rsim.sense(EBS_scene,EBS_unbal,Vvary = c(0,0), Dvary = c(0,0))
```

Each Ecosim parameter set can be subject to an initial simulation, also known as a 
burn-in period, to eliminate numerically unstable parameter sets. This instability 
usually results from incompatible draws of parameter sets that either lead to uncontrolled
population growth or population crash. Such as predator consumption and production 
that exceeds the production of prey. The length of the burn-in period can be set
in the original `rsim.scenario` object.

```{r , echo=TRUE}
# Setting the burn-in period in the EBS scenario object to 50 years.
EBS_scene$params$BURN_YEARS <- 50
```
During burn-in, if the biomass of a functional group exceeds 1,000x its starting
biomass or decreases to less than 1/1,000 its starting biomass, the ecosystem parameter
set is rejected and not retained for further analysis. A 50 year burn-in period is
generally sufficient to eliminate most of the unstable configurations but the length
of burn-in can be adjusted if necessary.

### Generating an ensemble of Ecosim parameter sets

To generate an ensemble of parameter sets we repeat the call to `rsim.sense` in 
a loop. We first determine the number of parameter sets we wish to generate
and create lists to store those parameter sets.
```{r , echo=TRUE}
NUM_RUNS <- 1000 # how many ecosystem parameter sets to generate
parlist<-as.list(rep(NA,NUM_RUNS)) # create lists to store the generated parameters
kept<-rep(NA,NUM_RUNS) # object to keep track of kept systems
set.seed(666) # Optional, set seed so output can be replicated
```
Now run the generator loop. In this example we will generate 1,000 parameter sets
(NUM_RUNS) for the eastern Bering Sea food web model. parlist[[i]] is the object
that will store the ith generated parameter set. In order to replicate the results
we use `set.seed`.

The following loop generates an ensemble of Ecosim parameters for the EBS food web 
model, assigns the generated base biomass (B_BaseRef) to the starting biomass in 
the `rsim.scenario` object, sets the burn years to 50, runs a simulation with `rsim.run`, 
and rejects or retains parameter sets based on the boundaries described above.

```{r generator loop, echo=TRUE, results='hide'}
for (i in 1:NUM_RUNS){
  EBSsense <- EBS_scene # scenario object
  # INSERT SENSE ROUTINE BELOW
  parlist[[i]]<- EBS_scene$params 		# Base ecosim params
  parlist[[i]]<- rsim.sense(EBS_scene,EBS_unbal,Vvary = c(-4.5,4.5), Dvary = c(0,0))	# Replace the base params with Ecosense params
  EBSsense$start_state$Biomass <- parlist[[i]]$B_BaseRef # Apply the Ecosense starting biomass
  parlist[[i]]$BURN_YEARS <- 50			# Set Burn Years to 50
  EBSsense$params <- parlist[[i]] # replace base params with the Ecosense generated params
  EBStest <- rsim.run(EBSsense, method="AB") # Run rsim with the generated system
  failList <- which(is.na(EBStest$end_state$Biomass))
  {if (length(failList)>0)
  {cat(i,": fail in year ",EBStest$crash_year,": ",EBStest$params$spname[failList],"\n"); kept[i]<-F; flush.console()}
    else 
    {cat(i,": success!\n"); kept[i]<-T;  flush.console()}} # output for the console
  parlist[[i]]$BURN_YEARS <- 1
}
```
This loop produces output in the console noting whether the generated parameter 
set was a success or failure, the year the ecosystem "crashed", and which group(s)
were responsible for the crash. We have suppressed that output here to save 
space but here is a sample of what that console output looks like:

  1 : fail in year  1 :  Salmon returning\
  2 : fail in year  1 :  Other zooplankton\
  3 : success!\
  4 : fail in year  1 :  Benthic amphipods\
  5 : fail in year  3 :  Polychaetes\

The first two ecosystems each failed in year one. Salmon returning was the group
that crashed in the first ecosystem and other zooplankton in the second. The third
ecosystem successfully survived the 50 year burn-in.

To determine which of the 1,000 generated systems survived burn-in, how many survived 
burn-in and what the rejection rate was:
```{r, echo=TRUE}
KEPT <- which(kept==T) # the number associated with the kept system
nkept <- length(KEPT); nkept # how many were kept
1-(nkept/NUM_RUNS) # rejection rate
```

Running each of the retained Ecosim parameter sets for 100 years without any additional 
perturbations, the first 50 years of which is the burn-in period.
```{r, echo=TRUE, results='hide'}
ecos <- as.list(rep(NA,length(KEPT))) # lists for simulated ecosystems
k <- 0  # counter for simulated ecosystems
for (i in KEPT) {
  EBS_scene$start_state$Biomass <- parlist[[i]]$B_BaseRef # set the starting Biomass to the generated values
  EBSsense <- EBS_scene # set up the scenario object
  EBSsense$params <- parlist[[i]] # set the params in the scenario object equal to the generated params.
  EBSsense$BURN_YEARS <- -1 # no burn-in period
  k <- k + 1 # set the number for the simulated ecosystem
  ecos[[k]] <- rsim.run(EBSsense,method='AB') # run rsim.run on the generated system
  print(c("Ecosystem no.", k, "out of", nkept)) # progress output to console
}
```

Calculate the relative biomass of each group in each system, relative to their starting
biomass as drawn by Ecosense.
```{r, echo=TRUE}
relB_ecos <- as.list(rep(NA,length(KEPT))) # list to output relative biomass
k <- 0
for (i in 1:nkept) {
  spname <- colnames(ecos[[i]]$out_Biomass[,2:ncol(ecos[[i]]$out_Biomass)])
  biomass <- ecos[[i]]$out_Biomass[, spname]
  n <- ncol(biomass)
  start.bio <- biomass[1, ] # the drawn starting biomass 
  start.bio[which(start.bio == 0)] <- 1
  rel.bio <- matrix(NA, dim(biomass)[1], dim(biomass)[2])
  for(isp in 1:n) rel.bio[, isp] <- biomass[, isp] / start.bio[isp] # biomass relative to biomass at t=1
  colnames(rel.bio) <- spname
  k <- k + 1
  relB_ecos[[k]] <- rel.bio
}
```
An example plot of walleye pollock. First setup a matrix with the pollock trajectories
from all of the retained parameter sets.
```{r, echo=TRUE}
this_species <- "Walleye pollock"
plot_mat <- matrix(nrow=1200, ncol=nkept) # matrix of pollock trajectories from all generated systems
for (i in 1:nkept) {
  plot_mat[,i] <- relB_ecos[[i]][,this_species]
}
```
Then plot all the trajectories:
```{r, echo=TRUE}
plot_col <- viridis(nkept)
layout(matrix(c(1,1,1,1,1,1,2,2), nrow = 1, ncol = 8, byrow = TRUE))
plot(1:1200, relB_ecos[[1]][,this_species], type='n', xlab="Months",
     ylab="Relative biomass", ylim=c(min(plot_mat),max(plot_mat)), main=this_species)
# one line for pollock in each of the generated systems
for (i in 1:nkept) {
  lines(1:1200, relB_ecos[[i]][,this_species], lwd=2, col=plot_col[i])
}
# distribution of pollock relative biomasses
boxplot(plot_mat[1200,], ylim=c(min(plot_mat),max(plot_mat)), yaxt='n')
axis(side=2, at=c(seq(0,150,50)), tick=TRUE, labels=F)
```

### Run the kept systems with a perturbation

Increase the FRate on pollock using `adjust.fishing`. The first 50 years are the 
burn-in period. So, the perturbation is put in place from year 51 to 100.
```{r, echo=TRUE, results='hide'}
ecos_sp <- as.list(rep(NA,length(KEPT))) # lists for simulated ecosystems
k <- 0  # counter for simulated ecosystems
for (i in KEPT) {
  EBS_scene$start_state$Biomass <- parlist[[i]]$B_BaseRef # set the starting Biomass to the generated values
  EBSsense <- EBS_scene # set up the scenario object
  EBSsense <- adjust.fishing(EBSsense, "ForcedFRate", group=this_species, sim.year=51:100, value=2) # perturb pollock FRate
  EBSsense$params <- parlist[[i]] # set the params in the scenario object equal to the generated params.
  EBSsense$BURN_YEARS <- -1 # no burn-in period
  k <- k + 1 # set the number for the simulated ecosystem
  ecos_sp[[k]] <- rsim.run(EBSsense,method='AB') # run rsim.run on the generated system
  print(c("Ecosystem no.", k, "out of", nkept)) # progress output to console
}
```
Calculate the biomass for each functional group relative to their biomass at the 
end of the burn-in period (year 50), for each of the retained ecosystems. 
```{r, echo=TRUE}
relB_ecos_sp <- as.list(rep(NA,length(KEPT)))
k <- 0
for (i in 1:nkept) {
  spname <- colnames(ecos_sp[[i]]$out_Biomass[,2:ncol(ecos_sp[[i]]$out_Biomass)])
  biomass <- ecos_sp[[i]]$out_Biomass[, spname]
  n <- ncol(biomass)
  start.bio <- biomass[600, ] # end of burn-in
  start.bio[which(start.bio == 0)] <- 1
  rel.bio <- matrix(NA, dim(biomass)[1], dim(biomass)[2])
  for(isp in 1:n) rel.bio[, isp] <- biomass[, isp] / start.bio[isp] # biomass relative to the end of burn-in biomass
  colnames(rel.bio) <- spname
  k <- k + 1
  relB_ecos_sp[[k]] <- rel.bio
}
```
Plot the trajectories for pollock from this perturbation. Note, the x-axis starts
at the begining of the perturbation (January of year 51).
```{r, echo=TRUE}
plot_mat_sp <- matrix(nrow=1200, ncol=nkept)
for (i in 1:nkept) {
  plot_mat_sp[,i] <- relB_ecos_sp[[i]][,this_species]
}
layout(matrix(c(1,1,1,1,1,1,2,2), nrow = 1, ncol = 8, byrow = TRUE))
plot(601:1200, relB_ecos_sp[[1]][601:1200,this_species], type='n', xlab="Months",
     ylab="Relative biomass", ylim=c(min(plot_mat_sp[601:1200,]),max(plot_mat_sp[601:1200,])), main=this_species)
for (i in 1:nkept) {
  lines(601:1200, relB_ecos_sp[[i]][601:1200,this_species], lwd=2, col=plot_col[i])
}
boxplot(plot_mat_sp[1200,], ylim=c(min(plot_mat_sp[601:1200,]),max(plot_mat_sp[601:1200,])))
```

A boxplot of the distribution of outcomes for all the living groups relative to 
their biomass at the start of the perturbation.
```{r, echo=TRUE}
sp_perturb_out <- matrix(nrow=nkept, ncol=(EBS_bal$NUM_LIVING+EBS_bal$NUM_DEAD))
for(i in 1:nkept){
  sp_perturb_out[i,] <- relB_ecos_sp[[i]][1200,]
}
colnames(sp_perturb_out) <- colnames(relB_ecos_sp[[1]])
par(mfrow=c(1,1), mar=c(9,3,1,1))
boxplot(sp_perturb_out, outline=FALSE, las=2, cex.axis=0.6)
abline(h=1, lty=2)
```
=======
---
title: "ecosense_proto_vignette"
author: "George A. Whitehouse"
date: "11/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rpath); library(data.table); library(viridis)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Rpath is an R implementation of the food web modeling program Ecopath with Ecosim
(EwE; Christensen and Pauly 1992^[Christensen V, Pauly D (1992) Ecopath II-a software 
for balancing steady-state ecosystem models and calculating network characteristics. 
Ecol Model 61:169-185. doi:10.1016/0304-3800(92)90016-8], Walters et al. 1997^[Walters 
C, Christensen V, Pauly D (1997) Structuring dynamic models of exploited ecosystems 
from trophic mass-balance assessments. Rev Fish Biol Fish 7:139-172. 
doi:10.1023/a:1018479526149]). For full documentation of the Rpath package see Lucey 
et al. (2020^[Lucey SM, Gaichas SK, Aydin KY (2020) Conducting reproducible ecosystem 
modeling using the open source mass balance model Rpath. Ecol Model 427:11. 
doi:10.1016/j.ecolmodel.2020.109057]). Ecosense is a Monte Carlo approach to generating 
an ensemble of plausible ecosystem parameter sets from a single Rpath model for 
use with `rsim.run`, the time-dynamic counterpart to Rpath. For complete documentation
of Ecosense see Whitehouse and Aydin (2020^[Whitehouse GA, Aydin KY (2020) 
Assessing the sensitivity of three Alaska marine food webs to perturbations: an 
example of Ecosim simulations using Rpath. Ecol Model 429:16. 
doi:10.1016/j.ecolmodel.2020.109074]). This vignette assumes the reader is familiar
with EwE and the Rpath package and its basic operations.

## Ecosense

To use Ecosense you need an `rpath.params` object and the corresponding `rsim.scenario`
object. In this vignette we've included the example Rpath models from Whitehouse and 
Aydin (2020) for the eastern Bering Sea (EBS), eastern Chukchi Sea (ECS), and Gulf 
of Alaska (GOA) marine food webs, and use the EBS model in our examples. Each of 
these models has 50–54 functional groups, including one fishery, one primary 
producer, two detrital compartments, and there are no stanzas.

```{r load the models and Ecosense}
load("data/sense_vignette.RData")
ls()
source("R/ecosense.R", local = knitr::knit_global())
```
Each of the three Rpath models included in this data set has an `rpath.params` object (unbal), 
a balanced `rpath` model (bal), and an `rsim.scenario` object (scene).

### Setting up Ecosense

Full sets of Ecosim parameter sets are drawn from distributions centered on the
original Rpath parameter estimates. The width of the distribution for each parameter
is defined in the data pedigree of the `rpath.params` object. The drawn parameters
include biomass, P/B, Q/B, diet composition, and natural mortality (i.e., M zero).
Additionally, vulnerability and handling time from the predator prey functional 
response can be drawn over a specified range (lower/upper bounds in log space-1) 
for each individual predator-prey interactions.

A single parameter set can be generated with a call to `rsim.sense`. The object 
returned by `rsim.sense` is equivalent to the list of parameters in an `rsim.scenario` 
object (e.g., EBS_scene$params).

```{r echo=TRUE}
# One set of Ecosim parameters for the EBS model
rsim.sense(EBS_scene,EBS_unbal,Vvary = c(0,0), Dvary = c(0,0))
```

Each Ecosim parameter set can be subject to an initial simulation, also known as a 
burn-in period, to eliminate numerically unstable parameter sets. This instability 
usually results from incompatible draws of parameter sets that either lead to uncontrolled
population growth or population crash. Such as predator consumption and production 
that exceeds the production of prey. The length of the burn-in period can be set
in the original `rsim.scenario` object.

```{r , echo=TRUE}
# Setting the burn-in period in the EBS scenario object to 50 years.
EBS_scene$params$BURN_YEARS <- 50
```
During burn-in, if the biomass of a functional group exceeds 1,000x its starting
biomass or decreases to less than 1/1,000 its starting biomass, the ecosystem parameter
set is rejected and not retained for further analysis. A 50 year burn-in period is
generally sufficient to eliminate most of the unstable configurations but the length
of burn-in can be adjusted if necessary.

### Generating an ensemble of Ecosim parameter sets

To generate an ensemble of parameter sets we repeat the call to `rsim.sense` in 
a loop. We first determine the number of parameter sets we wish to generate
and create lists to store those parameter sets.
```{r , echo=TRUE}
NUM_RUNS <- 1000 # how many ecosystem parameter sets to generate
parlist<-as.list(rep(NA,NUM_RUNS)) # create lists to store the generated parameters
kept<-rep(NA,NUM_RUNS) # object to keep track of kept systems
set.seed(666) # Optional, set seed so output can be replicated
```
Now run the generator loop. In this example we will generate 1,000 parameter sets
(NUM_RUNS) for the eastern Bering Sea food web model. parlist[[i]] is the object
that will store the ith generated parameter set. In order to replicate the results
we use `set.seed`.

The following loop generates an ensemble of Ecosim parameters for the EBS food web 
model, assigns the generated base biomass (B_BaseRef) to the starting biomass in 
the `rsim.scenario` object, sets the burn years to 50, runs a simulation with `rsim.run`, 
and rejects or retains parameter sets based on the boundaries described above.

```{r generator loop, echo=TRUE, results='hide'}
for (i in 1:NUM_RUNS){
  EBSsense <- EBS_scene # scenario object
  # INSERT SENSE ROUTINE BELOW
  parlist[[i]]<- EBS_scene$params 		# Base ecosim params
  parlist[[i]]<- rsim.sense(EBS_scene,EBS_unbal,Vvary = c(-4.5,4.5), Dvary = c(0,0))	# Replace the base params with Ecosense params
  EBSsense$start_state$Biomass <- parlist[[i]]$B_BaseRef # Apply the Ecosense starting biomass
  parlist[[i]]$BURN_YEARS <- 50			# Set Burn Years to 50
  EBSsense$params <- parlist[[i]] # replace base params with the Ecosense generated params
  EBStest <- rsim.run(EBSsense, method="AB") # Run rsim with the generated system
  failList <- which(is.na(EBStest$end_state$Biomass))
  {if (length(failList)>0)
  {cat(i,": fail in year ",EBStest$crash_year,": ",EBStest$params$spname[failList],"\n"); kept[i]<-F; flush.console()}
    else 
    {cat(i,": success!\n"); kept[i]<-T;  flush.console()}} # output for the console
  parlist[[i]]$BURN_YEARS <- 1
}
```
This loop produces output in the console noting whether the generated parameter 
set was a success or failure, the year the ecosystem "crashed", and which group(s)
were responsible for the crash. We have suppressed that output here to save 
space but here is a sample of what that console output looks like:

  1 : fail in year  1 :  Salmon returning\
  2 : fail in year  1 :  Other zooplankton\
  3 : success!\
  4 : fail in year  1 :  Benthic amphipods\
  5 : fail in year  3 :  Polychaetes\

The first two ecosystems each failed in year one. Salmon returning was the group
that crashed in the first ecosystem and other zooplankton in the second. The third
ecosystem successfully survived the 50 year burn-in.

To determine which of the 1,000 generated systems survived burn-in, how many survived 
burn-in and what the rejection rate was:
```{r, echo=TRUE}
KEPT <- which(kept==T) # the number associated with the kept system
nkept <- length(KEPT); nkept # how many were kept
1-(nkept/NUM_RUNS) # rejection rate
```

Running each of the retained Ecosim parameter sets for 100 years without any additional 
perturbations, the first 50 years of which is the burn-in period.
```{r, echo=TRUE, results='hide'}
ecos <- as.list(rep(NA,length(KEPT))) # lists for simulated ecosystems
k <- 0  # counter for simulated ecosystems
for (i in KEPT) {
  EBS_scene$start_state$Biomass <- parlist[[i]]$B_BaseRef # set the starting Biomass to the generated values
  EBSsense <- EBS_scene # set up the scenario object
  EBSsense$params <- parlist[[i]] # set the params in the scenario object equal to the generated params.
  EBSsense$BURN_YEARS <- -1 # no burn-in period
  k <- k + 1 # set the number for the simulated ecosystem
  ecos[[k]] <- rsim.run(EBSsense,method='AB') # run rsim.run on the generated system
  print(c("Ecosystem no.", k, "out of", nkept)) # progress output to console
}
```

Calculate the relative biomass of each group in each system, relative to their starting
biomass as drawn by Ecosense.
```{r, echo=TRUE}
relB_ecos <- as.list(rep(NA,length(KEPT))) # list to output relative biomass
k <- 0
for (i in 1:nkept) {
  spname <- colnames(ecos[[i]]$out_Biomass[,2:ncol(ecos[[i]]$out_Biomass)])
  biomass <- ecos[[i]]$out_Biomass[, spname]
  n <- ncol(biomass)
  start.bio <- biomass[1, ] # the drawn starting biomass 
  start.bio[which(start.bio == 0)] <- 1
  rel.bio <- matrix(NA, dim(biomass)[1], dim(biomass)[2])
  for(isp in 1:n) rel.bio[, isp] <- biomass[, isp] / start.bio[isp] # biomass relative to biomass at t=1
  colnames(rel.bio) <- spname
  k <- k + 1
  relB_ecos[[k]] <- rel.bio
}
```
An example plot of walleye pollock. First setup a matrix with the pollock trajectories
from all of the retained parameter sets.
```{r, echo=TRUE}
this_species <- "Walleye pollock"
plot_mat <- matrix(nrow=1200, ncol=nkept) # matrix of pollock trajectories from all generated systems
for (i in 1:nkept) {
  plot_mat[,i] <- relB_ecos[[i]][,this_species]
}
```
Then plot all the trajectories:
```{r, echo=TRUE}
plot_col <- viridis(nkept)
layout(matrix(c(1,1,1,1,1,1,2,2), nrow = 1, ncol = 8, byrow = TRUE))
plot(1:1200, relB_ecos[[1]][,this_species], type='n', xlab="Months",
     ylab="Relative biomass", ylim=c(min(plot_mat),max(plot_mat)), main=this_species)
# one line for pollock in each of the generated systems
for (i in 1:nkept) {
  lines(1:1200, relB_ecos[[i]][,this_species], lwd=2, col=plot_col[i])
}
# distribution of pollock relative biomasses
boxplot(plot_mat[1200,], ylim=c(min(plot_mat),max(plot_mat)), yaxt='n')
axis(side=2, at=c(seq(0,150,50)), tick=TRUE, labels=F)
```

### Run the kept systems with a perturbation

Increase the FRate on pollock using `adjust.fishing`. The first 50 years are the 
burn-in period. So, the perturbation is put in place from year 51 to 100.
```{r, echo=TRUE, results='hide'}
ecos_sp <- as.list(rep(NA,length(KEPT))) # lists for simulated ecosystems
k <- 0  # counter for simulated ecosystems
for (i in KEPT) {
  EBS_scene$start_state$Biomass <- parlist[[i]]$B_BaseRef # set the starting Biomass to the generated values
  EBSsense <- EBS_scene # set up the scenario object
  EBSsense <- adjust.fishing(EBSsense, "ForcedFRate", group=this_species, sim.year=51:100, value=2) # perturb pollock FRate
  EBSsense$params <- parlist[[i]] # set the params in the scenario object equal to the generated params.
  EBSsense$BURN_YEARS <- -1 # no burn-in period
  k <- k + 1 # set the number for the simulated ecosystem
  ecos_sp[[k]] <- rsim.run(EBSsense,method='AB') # run rsim.run on the generated system
  print(c("Ecosystem no.", k, "out of", nkept)) # progress output to console
}
```
Calculate the biomass for each functional group relative to their biomass at the 
end of the burn-in period (year 50), for each of the retained ecosystems. 
```{r, echo=TRUE}
relB_ecos_sp <- as.list(rep(NA,length(KEPT)))
k <- 0
for (i in 1:nkept) {
  spname <- colnames(ecos_sp[[i]]$out_Biomass[,2:ncol(ecos_sp[[i]]$out_Biomass)])
  biomass <- ecos_sp[[i]]$out_Biomass[, spname]
  n <- ncol(biomass)
  start.bio <- biomass[600, ] # end of burn-in
  start.bio[which(start.bio == 0)] <- 1
  rel.bio <- matrix(NA, dim(biomass)[1], dim(biomass)[2])
  for(isp in 1:n) rel.bio[, isp] <- biomass[, isp] / start.bio[isp] # biomass relative to the end of burn-in biomass
  colnames(rel.bio) <- spname
  k <- k + 1
  relB_ecos_sp[[k]] <- rel.bio
}
```
Plot the trajectories for pollock from this perturbation. Note, the x-axis starts
at the begining of the perturbation (January of year 51).
```{r, echo=TRUE}
plot_mat_sp <- matrix(nrow=1200, ncol=nkept)
for (i in 1:nkept) {
  plot_mat_sp[,i] <- relB_ecos_sp[[i]][,this_species]
}
layout(matrix(c(1,1,1,1,1,1,2,2), nrow = 1, ncol = 8, byrow = TRUE))
plot(601:1200, relB_ecos_sp[[1]][601:1200,this_species], type='n', xlab="Months",
     ylab="Relative biomass", ylim=c(min(plot_mat_sp[601:1200,]),max(plot_mat_sp[601:1200,])), main=this_species)
for (i in 1:nkept) {
  lines(601:1200, relB_ecos_sp[[i]][601:1200,this_species], lwd=2, col=plot_col[i])
}
boxplot(plot_mat_sp[1200,], ylim=c(min(plot_mat_sp[601:1200,]),max(plot_mat_sp[601:1200,])))
```

A boxplot of the distribution of outcomes for all the living groups relative to 
their biomass at the start of the perturbation.
```{r, echo=TRUE}
sp_perturb_out <- matrix(nrow=nkept, ncol=(EBS_bal$NUM_LIVING+EBS_bal$NUM_DEAD))
for(i in 1:nkept){
  sp_perturb_out[i,] <- relB_ecos_sp[[i]][1200,]
}
colnames(sp_perturb_out) <- colnames(relB_ecos_sp[[1]])
par(mfrow=c(1,1), mar=c(9,3,1,1))
boxplot(sp_perturb_out, outline=FALSE, las=2, cex.axis=0.6)
abline(h=1, lty=2)
```
>>>>>>> 0968e0d6581d08b23128e8dd3bc5ed0eef0adcb5
