---
title: "Rsim using R Ecosystem"
author: "Sean M. Lucey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Rpath is an implementation of the ecosystem model Ecopath with Ecosim (EwE; 
Christensen and Pauly 1992^[Christensen and Pauly. 1992. ECOPATH II - a software
for balancing steady-state models and calculating network characteristics. 
Ecological Modelling 61:169-85], Walters et al. 1997^[Walters et al. 1997. 
Structuring dynamic models of exploited ecosystems from trophic mass-balance 
assessments. Reviews of Fish Biology and Fisheries 7:1-34]).  This vignette 
describes some of the basic functionality of the package using a fictional 
ecosystem, R Ecosystem.  Any resemblance to an actual ecosystem is purely 
coincidental.  To see the underlying mathematics please refer to Lucey et al. 
(in prep^[Lucey et al. in prep. Improving the EBFM toolbox with an alternative 
open source version of Ecopath with Ecosim]).  

```{r rpath load, echo = F}
knitr::opts_chunk$set(
  comment = '#>',
  collapse = T)
library(Rpath); library(data.table)
````

## Running rsim

Rsim is the ecosim implementation of the EwE code in R.  In order to procede you
must have a valid Rpath object.  The steps for setting this up are discribed in 
the vignette "Rpath using R Ecosystem".

```{r creating Rpath object, echo = F}
#Groups and types for the R Ecosystem

groups <- c('Seabirds', 'Whales', 'Seals', 'JuvRoundfish1', 'AduRoundfish1', 
            'JuvRoundfish2', 'AduRoundfish2', 'JuvFlatfish1', 'AduFlatfish1',
            'JuvFlatfish2', 'AduFlatfish2', 'OtherGroundfish', 'Foragefish1',
            'Foragefish2', 'OtherForagefish', 'Megabenthos', 'Shellfish',
            'Macrobenthos', 'Zooplankton', 'Phytoplankton', 'Detritus', 
            'Discards', 'Trawlers', 'Midwater', 'Dredgers')

types  <- c(rep(0, 19), 1, rep(2, 2), rep(3, 3))

stgroups <- c(rep(NA, 3), rep('Roundfish1', 2), rep('Roundfish2', 2), 
              rep('Flatfish1', 2), rep('Flatfish2', 2), rep(NA, 14))

REco.params <- create.rpath.params(group = groups, type = types, stgroup = stgroups)

#Model
biomass <- c(0.0149, 0.454, NA, NA, 1.39, NA, 5.553, NA, 5.766, NA,
             0.739, 7.4, 5.1, 4.7, 5.1, NA, 7, 17.4, 23, 10, rep(NA, 5))

pb <- c(0.098, 0.031, 0.100, 2.026, 0.42, 2.1, 0.425, 1.5, 0.26, 1.1, 0.18, 0.6,
        0.61, 0.65, 1.5, 0.9, 1.3, 7, 39, 240, rep(NA, 5))

qb <- c(76.750, 6.976, 34.455, NA, 2.19, NA, 3.78, NA, 1.44, NA, 1.69,
        1.764, 3.52, 5.65, 3.6, 2.984, rep (NA, 9))

REco.params$model[, Biomass := biomass]
REco.params$model[, PB := pb]
REco.params$model[, QB := qb]

#EE for groups w/o biomass
REco.params$model[Group %in% c('Seals', 'Megabenthos'), EE := 0.8]

#Production to Consumption for those groups without a QB
REco.params$model[Group %in% c('Shellfish', 'Zooplankton'), ProdCons:= 0.25]
REco.params$model[Group == 'Macrobenthos', ProdCons := 0.35]

#Biomass accumulation and unassimilated production
REco.params$model[, BioAcc  := c(rep(0, 22), rep(NA, 3))]
REco.params$model[, Unassim := c(rep(0.2, 18), 0.4, rep(0, 3), rep(NA, 3))]

#Detrital Fate
REco.params$model[, Detritus := c(rep(1, 20), rep(0, 5))]
REco.params$model[, Discards := c(rep(0, 22), rep(1, 3))]

#Fisheries
#Landings
trawl  <- c(rep(0, 4), 0.08, 0, 0.32, 0, 0.09, 0, 0.05, 0.2, rep(0, 10), rep(NA, 3))
mid    <- c(rep(0, 12), 0.3, 0.08, 0.02, rep(0, 7), rep(NA, 3))
dredge <- c(rep(0, 15), 0.1, 0.5, rep(0, 5), rep(NA, 3))
REco.params$model[, Trawlers := trawl]
REco.params$model[, Midwater := mid]
REco.params$model[, Dredgers := dredge]

#Discards
trawl.d  <- c(1e-5, 1e-7, 0.001, 0.001, 0.005, 0.001, 0.009, 0.001, 0.04, 0.001,
              0.01, 0.08, 0.001, 0.001, 0.001, rep(0, 7), rep(NA, 3))
mid.d    <- c(rep(0, 2), 0.001, 0.001, 0.01, 0.001, 0.01, rep(0, 4), 0.05, 0.05,
              0.01, 0.01, rep(0, 7), rep(NA, 3))
dredge.d <- c(rep(0, 3), 0.001, 0.05, 0.001, 0.05, 0.001, 0.05, 0.001, 0.01, 0.05,
              rep(0, 3), 0.09, 0.01, 1e-4, rep(0, 4), rep(NA, 3))
REco.params$model[, Trawlers.disc := trawl.d]
REco.params$model[, Midwater.disc := mid.d]
REco.params$model[, Dredgers.disc := dredge.d]

#Group parameters
REco.params$stanzas$stgroups[, VBGF_Ksp := c(0.145, 0.295, 0.0761, 0.112)]
REco.params$stanzas$stgroups[, Wmat     := c(0.0769, 0.561, 0.117,  0.321)]

#Individual stanza parameters
REco.params$stanzas$stindiv[, First   := c(rep(c(0, 24), 3), 0, 48)]
REco.params$stanzas$stindiv[, Last    := c(rep(c(23, 400), 3), 47, 400)]
REco.params$stanzas$stindiv[, Z       := c(2.026, 0.42, 2.1, 0.425, 1.5, 
                                           0.26, 1.1, 0.18)]
REco.params$stanzas$stindiv[, Leading := rep(c(F, T), 4)]

REco.params <- rpath.stanzas(REco.params)

#Diets
REco.params$diet[, Seabirds        := c(rep(NA, 11), 0.1, 0.25, 0.2, 0.15, 
                                         rep(NA, 6), 0.3, NA)]
REco.params$diet[, Whales          := c(rep(NA, 3), 0.01, NA, 0.01, NA, 0.01, 
                                         NA, 0.01, rep(NA, 4), 0.1, rep(NA, 3), 
                                         0.86, rep(NA, 4))]
REco.params$diet[, Seals           := c(rep(NA, 3), 0.05, 0.1, 0.05, 0.2, 0.005, 
                                         0.05, 0.005, 0.01, 0.24, rep(0.05, 4), 
                                         0.09, rep(NA, 6))]
REco.params$diet[, JuvRoundfish1   := c(rep(NA, 3), rep(c(1e-4, NA), 4), 1e-3, 
                                         rep(NA, 2), 0.05, 1e-4, NA, .02, 0.7785, 
                                         0.1, 0.05, NA, NA)]
REco.params$diet[, AduRoundfish1   := c(rep(NA, 5), 1e-3, 0.01, 1e-3, 0.05, 1e-3, 
                                         0.01, 0.29, 0.1, 0.1, 0.347, 0.03, NA, 
                                         0.05, 0.01, rep(NA, 4))]
REco.params$diet[, JuvRoundfish2   := c(rep(NA, 3), rep(c(1e-4, NA), 4), 1e-3, 
                                         rep(NA, 2), 0.05, 1e-4, NA, .02, 0.7785, 
                                         0.1, .05, NA, NA)]
REco.params$diet[, AduRoundfish2   := c(rep(NA, 3), 1e-4, NA, 1e-4, NA, rep(1e-4, 4), 
                                         0.1, rep(0.05, 3), 0.2684, 0.01, 0.37, 0.001, 
                                         NA, 0.1, NA, NA)]
REco.params$diet[, JuvFlatfish1    := c(rep(NA, 3), rep(c(1e-4, NA), 4), rep(NA, 3), 
                                         rep(1e-4, 2), NA, 0.416, 0.4334, 0.1, 0.05, 
                                         NA, NA)]
REco.params$diet[, AduFlatfish1    := c(rep(NA, 7), rep(1e-4, 5), rep(NA, 2), 0.001, 
                                         0.05, 0.001, 0.6, 0.2475, NA, 0.1, NA, NA)]
REco.params$diet[, JuvFlatfish2    := c(rep(NA, 3), rep(c(1e-4, NA), 4), rep(NA, 3),
                                         rep(1e-4, 2), NA, 0.416, 0.4334, 0.1, 0.05, 
                                         NA, NA)]
REco.params$diet[, AduFlatfish2    := c(rep(NA, 7), 1e-4, NA, 1e-4, rep(NA, 4), 
                                         rep(1e-4, 3), 0.44, 0.3895, NA, 0.17, NA, NA)]
REco.params$diet[, OtherGroundfish := c(rep(NA, 3), rep(1e-4, 8), 0.05, 0.08, 0.0992, 
                                         0.3, 0.15, 0.01, 0.3, 0.01, rep(NA, 4))]
REco.params$diet[, Foragefish1     := c(rep(NA, 3), rep(c(1e-4, NA), 4), rep(NA, 7), 
                                         0.8196, 0.06, 0.12, NA, NA)]
REco.params$diet[, Foragefish2     := c(rep(NA, 3), rep(c(1e-4, NA), 4), rep(NA, 7), 
                                         0.8196, 0.06, 0.12, NA, NA)]
REco.params$diet[, OtherForagefish := c(rep(NA, 3), rep(c(1e-4, NA), 4), rep(NA, 7), 
                                         0.8196, 0.06, 0.12, NA, NA)]
REco.params$diet[, Megabenthos     := c(rep(NA, 15), 0.1, 0.03, 0.55, rep(NA, 2), 0.32,
                                         NA, NA)]
REco.params$diet[, Shellfish       := c(rep(NA, 18), 0.3, 0.5, 0.2, NA, NA)]
REco.params$diet[, Macrobenthos    := c(rep(NA, 16), 0.01, rep(0.2, 2), NA, 0.59, NA, NA)]
REco.params$diet[, Zooplankton     := c(rep(NA, 18), 0.2, 0.6, 0.2, NA, NA)]

REco <- rpath(REco.params, eco.name = 'R Ecosystem')
```

Running rsim, is a three part process.  First, the function `rsim.scenario` is run
to convert rpath parameters to rates.  Within `rsim.scenario` are 5 functions that 
initialize the basic and stanza parameters, creates perturbation matrices for fishing 
and other forcing functions, and a list of initial states.  Arguments passed to the 
`rsim.scenario` function are the Rpath object, the rpath parameter object, and a vector 
of years corresponding to the length of the simulation.

```{r rsim.scenario}
REco.sim <- rsim.scenario(REco, REco.params, years = 1:100)
```

The second part of rsim is to add forcing functions or change the fishing behavior.
This is accomplished by changing the appropriate list within the Rsim.scenario object
created in the first step.  There are a series of adjust functions that will do this
without having to know the specific group numbers.  For example, we can double the
effort of the trawler fleet after 25 years using the `adjust.fishing` function.

```{r Change effort}
REco.sim <- adjust.fishing(REco.sim, 'EFFORT', group = 'Trawlers', sim.year = 25:100, 
                           value = 2)
```

The final part is to apply the modified Rsim.scenario function to the `rsim.run`
function.  The only arguments to the `rsim.run` function are the Rsim.scenario
object, the method for numerical integration, and the length of the simulation.
Rpath allows for both Adams-Bashforth and Runge-Kutta 4 numerical integration.
Older versions of EwE use the Adams-Bashforth method while the latest version (6+)
uses Runge-Kutta (The default method for Rpath).

```{r rsim.run}
REco.run1 <- rsim.run(REco.sim, method = 'RK4', years = 1:100)
```

The output from `rsim.run` is another S3 object class called 'Rsim.output'.  Similar
to the Rpath object created by `rpath`, the generic function `print` will give
output similar to the 'Ecosim results' tab in EwE. If you want to save the `print`
results you need to use the function `write.rpath.sim`.  The function `summary` just
like for the Rpath object will display what other list items are available.

Add code for rsim print...

In addition, there is a quick graphical routine for plotting
biomass trajectories over time.  Other plots similar to the group plots are possible
but at this time we do not have a built-in function.

```{r ecosim equilibrium plot, fig.height = 7, fig.width = 9}
rsim.plot(REco.run1, groups[1:22])
```


