---
title: "Run a dynamic food web simulation in Rsim"
author: "Sean M. Lucey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run a dynamic food web simulation in Rsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r rpath load, echo = F}
knitr::opts_chunk$set(
  comment = '#>',
  collapse = T)
library(Rpath); library(data.table)
```

Rsim is the ecosim implementation of the EwE code in R.  In order to proceed you must have a valid Rpath object.  The steps for setting this up are described in `vignette("ModelSetup")`.

```{r setup, echo = TRUE, eval =TRUE}
REco <- rpath(REco.params, eco.name = 'R Ecosystem')
```

Running rsim, is a three part process.  First, the function `rsim.scenario()` is run to convert rpath parameters to rates.  Within `rsim.scenario()` are 5 functions that initialize the basic and stanza parameters, creates perturbation matrices for fishing 
and other forcing functions, and a list of initial states.  Arguments passed to the `rsim.scenario()` function are the Rpath object, the rpath parameter object, and a vector of years corresponding to the length of the simulation.

```{r rsim.scenario, echo = TRUE, eval = TRUE}
REco.sim <- rsim.scenario(REco, REco.params, years = 1:100)
```

The second part of rsim is to add forcing functions or change the fishing behavior. This is accomplished by changing the appropriate list within the `REco.sim` object created in the first step. There are a series of adjust functions that will do this without having to know the specific group numbers.  For example, we can double the effort of the trawler fleet after 25 years using the `adjust.fishing()` function.

```{r Change effort, echo = TRUE, eval = TRUE}
REco.sim <- adjust.fishing(REco.sim, 'ForcedEffort', group = 'Trawlers', sim.year = 25:100, 
                           value = 2)
```

The final part is to pass the modified REco.sim object to the `rsim.run()` function.  The only arguments to the `rsim.run()` function are the Rco.sim object, the method for numerical integration, and the length of the simulation. Rpath allows for both Adams-Bashforth and Runge-Kutta 4 numerical integration. Older versions of EwE use the Adams-Bashforth method while the latest version (6+) uses Runge-Kutta (The default method for Rpath).

```{r rsim.run}
REco.run1 <- rsim.run(REco.sim, method = 'RK4', years = 1:100)
```

The output from `rsim.run()` is another S3 class object. Similar to the Rpath object created by `rpath()`, the generic function `print` will give output similar to the 'Ecosim results' tab in EwE. If you want to save the `print` results you need to use the function `write.Rsim()`. The function `summary` just like for the Rpath object will display what other list items are available.

For example:

```{r print_runrsim, echo = TRUE, eval = TRUE}
print(REco.run1)
```

To export this object to a csv file 

```{r write_runrsim, echo = TRUE, eval = FALSE}
write.Rsim(REco.run1,file = here::here("run1Output.csv"))
```

```{r summary_runrsim, echo = TRUE, eval = TRUE}
summary(REco.run1)
```

In addition, there is a quick graphical routine for plotting
biomass trajectories over time.  Other plots similar to the group plots are possible but at this time there isn't a built-in function.

```{r ecosimEquilibriumPlot, fig.height = 7, fig.width = 9,fig.alt="Biomass Trajectories of all species in model"}
rsim.plot(REco.run1, spname="all")
```


