---
title: "Create a static food web model in Rpath"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create a static food web model in Rpath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(Rpath); library(data.table)
```

## Running Rpath

After following the Model Setup steps (`vignette("ModelSetup")`) and creating the parameter object, balancing the model in Rpath is relatively straightforward. Simply pass the parameter object as an argument to the `rpath()` function. If desired, an ecosystem name (`eco.name`) and the area (`eco.area`, in km^2) of the ecosystem can be attributed to the model.

```{r Running ecopath, echo = T, eval = T}
REco <- rpath(REco.params, eco.name = 'R Ecosystem')
REco
```

The output object from `rpath()` is an S3 object type called 'Rpath'.  Rpath objects are a list of parameters from the mass balance.  However, the `print` function will display the same information as the "Basic Estimates" tab from EwE. You will also notice that the `print` function will display whether the model is balanced or not.

If the model was not balanced, it would list the groups that are not balanced.

You can also display the mortalities associated with each group by supplying the argument `morts = T` to the `print` function.

```{r Ecopath morts, echo = T, eval = T}
print(REco, morts = T) 
```

Note: If you wish to save the `print` output (to a .csv or .RData file) you need to use the function `write.rpath()`. This function will also accept the argument `morts = T`.

The generic function `summary` will display some summary statistics on the model as well as a list of attributes you can access.

```{r Ecopath summaries}
summary(REco)

```

To access any of the other attributes simply use the standard list notation. For example, to view the trophic level information

```{r Ecopath summariesInd, eval = T, echo = T}
REco$TL
```

One of the advantages of R is its graphical ability.  Users can feel free to develop their own graphical routines for the Rpath outputs.  However, we have included a basic food web plot.  The routine can include fisheries, display group numbers or names, and even highlight a particular group.

```{r FoodWebPlot1, fig.align = 'center', fig.height = 7, fig.width = 7, fig.alt="Food web of the example R ecosystem"}
webplot(REco)
```


```{r FoodWebPlot2, fig.align = 'center', fig.height = 7, fig.width = 7, fig.alt="Food web with vertex labels of the example R ecosystem"}
webplot(REco, labels = T)
```

```{r FoodWebPlot3, fig.align = 'center', fig.height = 7, fig.width = 7,, fig.alt="Food web of the xample R ecosystem with fishing fleets included and AduRoundfish1 highlighted"}
webplot(REco, fleets = T, highlight = 'AduRoundfish1')
```

